---
title: "post1_analysis"
output: html_document
date: "2025-06-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(ggpubr)
library(Seurat)
library(dplyr)
library(tibble)
library(ggplot2)
library(patchwork)
library(org.Dr.eg.db)
library(pheatmap)
library(clusterProfiler)
library(devtools)
library(speckle)
library(DESeq2)
```

```{r}

mydata <- readRDS("/gpfs/home/malm/bgmp/bioinfo/scRNAseq-ZFish/all.treatments.integrated.ann_3.rds")
```


### Defining EEC cells using Daniocell gene list 



```{r}
DefaultAssay(mydata) <- "RNA"

eec_genes <- c("adcyap1a", "penka", "nmbb", "urah", "adgrg4b", "rxpf3.3b", 
               "tmem176l.2", "alpk3b", "adgrg4a", "ugt5c3", "hrh2b", "lmx1ba",
               "slc7a14a", "tspan34", "lrmp", "tmem196b", "galr1a", "plac8.2",
               "cacna2d4a", "kcnk13a", "cldn15la", "nkx2.2a", "zak", "kcnk3b",
               "mapk15", "zdhhc15b", "gpa33b", "XLOC-013854", "ampd2b", "malrd1")

# Convert to Ensembl IDs
eec_ensembl <- mapIds(
  org.Dr.eg.db,
  keys = eec_genes,
  keytype = "SYMBOL",
  column = "ENSEMBL",
  multiVals = "first"
)

# Remove NA values (genes that didn't map)
eec_ensembl <- na.omit(eec_ensembl)
eec_ensembl <- unname(eec_ensembl) 
```


### How many genes mapped to our dataset? What are the missing genes? 


```{r, message=FALSE}
existing_genes <- eec_ensembl[eec_ensembl %in% rownames(mydata)]
print(paste(length(existing_genes), "out of", length(eec_ensembl), "genes found"))


```


```{r, message=FALSE,fig.width=10, fig.height=8}

mydata <- AddModuleScore(
  object = mydata,
  features = list(EEC_signature = existing_genes),  
  name = "EEC_",
  ctrl = 100,  
  seed = 42 
)

threshold <- quantile(mydata$EEC_1, probs = 0.95)
mydata$EEC_status <- ifelse(mydata$EEC_1 > threshold, "EEC", "Non-EEC")

# Visualize
DimPlot(mydata, group.by = "EEC_status", cols = c("red", "grey"))


#VlnPlot(mydata, features = "EEC_1", group.by = "seurat_clusters")

```


### How many cells were marked as EEC? 


```{r}

table(mydata@meta.data$EEC_status)
```

## Enrichment analysis of newly identified EEC cells 

```{r, warning=FALSE}

eec_cells <- subset(mydata, EEC_status == "EEC")


deg_EEC <- FindMarkers(eec_cells,
                      group.by = "treatment",
                      ident.1 = "befa",
                      ident.2 = "gf",
                      logfc.threshold = 0.25,
                      min.pct = 0.1)


deg_genes <- rownames(deg_EEC[deg_EEC$p_val_adj < 0.05 & abs(deg_EEC$avg_log2FC) > 0.5, ])


background_genes <- rownames(eec_cells)[rowSums(GetAssayData(eec_cells)) > 0]


entrez_ids <- mapIds(org.Dr.eg.db,
                    keys = deg_genes,
                    column = "ENTREZID",
                    keytype = "ENSEMBL",
                    multiVals = "first") %>%
             na.omit()


ego2 <- enrichGO(gene = entrez_ids,
                universe = names(background_genes),  # All possible genes
                OrgDb = org.Dr.eg.db,
                ont = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                qvalueCutoff = 0.1)


dotplot(simplify(ego2)) + 
  ggtitle("befa vs gf in EEC Cells")
```

### Subsetting to clusters labeled as intestine (17,30,47)

```{r}

target_clusters <- subset(mydata, idents = c(17, 30, 47))


Idents(target_clusters) <- "treatment"
deg_results <- FindMarkers(
  target_clusters,
  ident.1 = "befa",
  ident.2 = "gf",
  logfc.threshold = 0.25,
  min.pct = 0.1
)

# 3. Filter significant genes
deg_geness <- rownames(deg_results[deg_results$p_val_adj < 0.1 & abs(deg_results$avg_log2FC) > 0.5, ])

# 4. Prepare for GO enrichment
background_geness <- rownames(target_clusters)[rowSums(GetAssayData(target_clusters)) > 0]
entrez_idss <- mapIds(
  org.Dr.eg.db,
  keys = deg_geness,
  column = "ENTREZID",
  keytype = "ENSEMBL",
  multiVals = "first"
) %>% na.omit()

# 5. Run GO enrichment
# ego <- enrichGO(
#   gene = entrez_idss,
#   universe = background_geness,
#   OrgDb = org.Dr.eg.db,
#   ont = "BP",
#   pAdjustMethod = "BH",
#   pvalueCutoff = 0.1
# )


eggo <- enrichKEGG(gene = entrez_idss, organism = "dre")
# 6. Plot enriched pathways
dotplot(eggo) + 
  ggtitle("BEFA vs GF in Clusters 17,30,47")

```


## Cluster-Level Treatment Effects

### Are there sepecefic cell types that are more sensitive to befa treatment compared to other cell types? 


```{r, message=FALSE}

results <- data.frame()

# Iterate through all clusters
for (cluster in levels(mydata$seurat_clusters)) {
  cluster_cells <- subset(mydata, idents = cluster)
  Idents(cluster_cells) <- "treatment"
  

  deg <- FindMarkers(
    cluster_cells,
    ident.1 = "befa",
    ident.2 = "gf",
    min.pct = 0.1,
    logfc.threshold = 0.25
  )
  

  results <- rbind(results, data.frame(
    Cluster = cluster,
    n_Upregulated = sum(deg$avg_log2FC > 0 & deg$p_val_adj < 0.05),
    n_Downregulated = sum(deg$avg_log2FC < 0 & deg$p_val_adj < 0.05),
    Median_logFC = median(deg$avg_log2FC),
    stringsAsFactors = FALSE
  ))
}
```


Finding sensitive clusters (transcriptionally responsive):

    For each cluster, compared gene expression between treatment groups ("befa" vs "gf").

Only genes with:

        >25% expression change (log2FC > 0.25).

        Statistical significance (adjusted *p* < 0.05).

Scoring Clusters:

    Total DEGs: Number of significantly up/downregulated genes.

    Effect Size: Median magnitude of expression changes (abs(log2FC)).

    Sensitivity Score: (DEGs per 100 cells) × (Median Effect Size).

        Normalized by cluster size to avoid bias (e.g., large clusters dominating).

Selection Criteria:

    Minimum Activity: ≥10 DEGs per cluster.

    Above-Average Impact: Sensitivity Score > median across all clusters.

    Quality Control: Excluded clusters with <10 cells in either treatment group


```{r, warning=FALSE}
# Add cell counts per treatment to results
results$n_cells_befa <- table(mydata$seurat_clusters[mydata$treatment == "befa"])[results$Cluster]
results$n_cells_gf <- table(mydata$seurat_clusters[mydata$treatment == "gf"])[results$Cluster]

# Calculate minimum cells per treatment (conservative normalization)
results$n_cells_min <- pmin(results$n_cells_befa, results$n_cells_gf, na.rm=TRUE)

# Compute size-adjusted metrics
results$DEGs_per_100_cells <- ifelse(
  results$n_cells_min > 0, 
  (results$Total_DEGs / results$n_cells_min) * 100, 
  0  # Avoid division by zero
)
results$Sensitivity_Score <- results$DEGs_per_100_cells * abs(results$Median_logFC)

# Filter for sensitive clusters (≥10 DEGs & above-median score)
sensitive_clusters <- results[
  results$Total_DEGs >= 10 & 
  results$Sensitivity_Score > median(results$Sensitivity_Score, na.rm=TRUE),
]
sensitive_clusters <- sensitive_clusters[order(-sensitive_clusters$Sensitivity_Score), ]
    
```



## How are muscle cells impacted ? 


```{r}
# Get all muscle cells
muscle_cells <- subset(mydata, 
                      subset = grepl("MusSkelFast|MusSkelSlow", singleR.labels_miller))

# Verify cell counts
#table(muscle_cells$singleR.labels_miller)

Idents(muscle_cells) <- "treatment"
muscle_markers <- FindMarkers(
  muscle_cells,
  ident.1 = "befa",
  ident.2 = "gf",
  min.pct = 0.25,
  logfc.threshold = 0.1,
  only.pos = TRUE,
  random.seed = 67,
  test.use = "wilcox"
)


sig_markers <- subset(muscle_markers, p_val_adj < 0.1 & abs(avg_log2FC) > 1)
```


```{r, message=FALSE, fig.width=8, fig.height=12}

top_degs <- head(rownames(sig_markers), 12)


gene_symbols <- mapIds(
  org.Dr.eg.db,
  keys = top_degs[1:12],       
  column = "SYMBOL",
  keytype = "ENSEMBL"
)



plot_list <- lapply(1:11, function(i) {
  VlnPlot(
    muscle_cells,
    features = top_degs[i],
    split.by = "treatment",
    pt.size = 0
  ) +
    ggtitle(gene_symbols[i]) +  
    scale_fill_manual(values = c("befa" = "red3", "gf" = "steelblue"))
})

# using patchwork

wrap_plots(plot_list, ncol = 3)

```



## Impact on Nerural cells 


```{r}

neural_cell_types <- c("RetNeuron5", "RadialGlia5", "RetPR", 
                       "NCcranial5a", "NCcranial5b", "NCcranial5e",
                       "NCcart5a", "NCcart5b", "NCmel25", "NCirid",
                       "NCxantho25", "IonoNCC5a", "IonoNCC5b", 
                       "Epiphysis", "HairCell")

neural_cells <- subset(mydata, 
                       subset = singleR.labels_miller %in% neural_cell_types)

# Verify cell counts
#table(muscle_cells$singleR.labels_miller)

Idents(neural_cells) <- "treatment"
neural_markers <- FindMarkers(
  neural_cells,
  ident.1 = "befa",
  ident.2 = "gf",
  min.pct = 0.25,
  logfc.threshold = 0.1,
  only.pos = TRUE,
  test.use = "wilcox"
)


sig_markers_neural <- subset(neural_markers, p_val_adj < 0.1 & abs(avg_log2FC) > 1)

```


```{r, message=FALSE, fig.width= 8, fig.height=12}

top_degs_neural <- head(rownames(sig_markers_neural), 12)

neural_symbols <- mapIds(
  org.Dr.eg.db,
  keys = top_degs_neural[1:12],       
  column = "SYMBOL",
  keytype = "ENSEMBL"
)

plot_list_n <- lapply(1:11, function(i) {
  VlnPlot(
    neural_cells,
    features = top_degs_neural[i],
    split.by = "treatment",
    pt.size = 0
  ) +
    ggtitle(neural_symbols[i]) +  
    scale_fill_manual(values = c("befa" = "red3", "gf" = "steelblue"))
})



wrap_plots(plot_list_n, ncol = 3)

```


## Top enriched pathways across all clusters (befa vs gf)

```{r, message=FALSE}
# Subset for befa and gf
mydata_subset <- subset(mydata, subset = treatment %in% c("befa", "gf"))



Idents(mydata_subset) <- "treatment"

mydata_subset <- SetIdent(mydata_subset, value = factor(mydata_subset$treatment, levels = c("gf", "befa")))

de_results <- FindMarkers(
  object = mydata_subset,
  ident.1 = "befa",  
  ident.2 = "gf",    
  test.use = "MAST",
  logfc.threshold = 0.25, 
  min.pct = 0.1,
  verbose = TRUE
)
```

```{r}

de_results$gene <- rownames(de_results)
de_results$entrez_id <- mapIds(
  org.Dr.eg.db,
  keys = rownames(de_results),
  column = "ENTREZID",
  keytype = "ENSEMBL",
  multiVals = "first"
)

de_results_filtered <- de_results[!is.na(de_results$entrez_id) & de_results$p_val_adj < 0.05, ]

```


```{r}


up_genes <- de_results_filtered$entrez_id[de_results_filtered$avg_log2FC > 0]


down_genes <- de_results_filtered$entrez_id[de_results_filtered$avg_log2FC < 0]


up_go <- enrichGO(up_genes, org.Dr.eg.db, ont = "BP")
down_go <- enrichGO(down_genes, org.Dr.eg.db, ont = "BP")
```

```{r}

dotplot(up_go, showCategory = 10, title = "Upregulated Pathways (befa vs. gf)")

dotplot(down_go, showCategory = 10, title = "Downregulated Pathways (befa vs. gf)")
```


### What genes are driving each of the pathways 

```{r, message=FALSE}
# For upregulated GO terms
up_go_top_genes <- lapply(up_go@result$geneID[1:5], function(term_genes) {
  genes <- unlist(strsplit(term_genes, "/"))
  head(genes, 20)  
}) %>% unlist() %>% unique()

# For downregulated GO terms
down_go_top_genes <- lapply(down_go@result$geneID[1:5], function(term_genes) {
  genes <- unlist(strsplit(term_genes, "/"))
  head(genes, 20)  
}) %>% unlist() %>% unique()


gene_annotations <- mapIds(
  org.Dr.eg.db,
  keys = rownames(mydata_subset),  
  column = "SYMBOL",               
  keytype = "ENSEMBL",            
  multiVals = "first"             
)

# Add symbols to your Seurat object
#mydata_subset[[ "RNA" ]]@meta.features$symbol <- gene_annotations


# For upregulated GO genes
up_go_ensembl <- mapIds(
  org.Dr.eg.db,
  keys = up_go_top_genes,  
  column = "ENSEMBL",      
  keytype = "ENTREZID"
) %>% na.omit()

# For downregulated GO genes
down_go_ensembl <- mapIds(
  org.Dr.eg.db,
  keys = down_go_top_genes,
  column = "ENSEMBL",
  keytype = "ENTREZID"
) %>% na.omit()


```



```{r, fig.width= 12, fig.height=8}
# Combine all genes of interest
top_ensembl <- unique(c(up_go_ensembl, down_go_ensembl))
top_symbols <- mapIds(
  org.Dr.eg.db,
  keys = top_ensembl,
  keytype = "ENSEMBL",
  column = "SYMBOL"
)
  


# Create labeled DotPlot
DotPlot(
  mydata,
  features = top_ensembl,  
  group.by = "treatment",
  cols = c("blue", "red"),
  dot.scale = 6
) +
  scale_x_discrete(
    labels = top_symbols  
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.title.x = element_blank()
  ) +
  labs(title = "Top GO-Driven Lipid Genes: befa vs. gf")
```

## Impact of befa on lipid metabolism 



```{r}
# Subset cells for BEFA and GF groups
bafa_cells <- subset(mydata, subset = treatment == "befa")
gf_cells <- subset(mydata, subset = treatment == "gf")

```



```{r}

de_lip <- FindMarkers(
  mydata,
  ident.1 = "befa",
  ident.2 = "gf",
  group.by = "treatment",
  test.use = "wilcox",
  logfc.threshold = 0.25,
  min.pct = 0.1,
  only.pos = FALSE
)

# Filter for significant genes (FDR < 0.05)
sig_lip <- de_lip %>% 
  filter(p_val_adj < 0.05) %>% 
  arrange(desc(avg_log2FC))


```

```{r, message=FALSE}
lipid_gene_symbols <- c(
  
  # PPAR/FAO pathway
  
  "acox1", "cpt1a", "acadvl",  
  # Lipogenesis
  
  "fasn", "acaca", "elovl6",  
  
  # Lipolysis/Lipid transport
  
  "lipc", "apoa1", "apoe", "fabp1a", "fabp10", 
  
  # Cholesterol/Steroid
  
  "hmgcr", "srebf2", "cyp51", "dhcr7"
)


lipid_genes_ensembl <- mapIds(
  org.Dr.eg.db,
  keys = lipid_gene_symbols,
  keytype = "SYMBOL",
  column = "ENSEMBL"
) %>% na.omit()
```


```{r, fig.width= 12, fig.height=8, warning=FALSE}
mydata_subset1 <- subset(mydata, subset = treatment %in% c("gf", "befa"))


DotPlot(
  mydata_subset1,
  features = lipid_genes_ensembl,
  group.by = "treatment",          
  cols = c("blue", "red"),        
  dot.scale = 6
) +
  #scale_x_discrete(labels = gene_labels) +  
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    #axis.title.x = element_blank(), 
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    title = "Lipid Metabolism Genes: BEFA vs. GF",
    y = "Treatment" 
  ) +
  guides( 
    color = guide_colorbar(title = "Avg. Expression"),
    size = guide_legend(title = "% Expressed")
)
```
```{r}

go_bp <- enrichGO(
  gene          = lipid_genes_ensembl,
  OrgDb         = org.Dr.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",             
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.1,
  readable      = TRUE                
)

# Filter for lipid-related terms
lipid_go_bp <- go_bp %>% 
  filter(grepl("lipid|fatty acid|sterol|cholesterol|glycerolipid", Description, ignore.case = TRUE))

barplot(lipid_go_bp, 
        showCategory = 10, 
        title = "GO-BP: Lipid-Related Pathways",
        font.size = 10)

```

```{r}

go_mf <- enrichGO(
  gene          = lipid_genes_ensembl,
  OrgDb         = org.Dr.eg.db,
  keyType       = "ENSEMBL",
  ont           = "MF",               # Molecular Function
  pvalueCutoff  = 0.05
)

dotplot(lipid_go_bp, 
        title = "Enriched Lipid Pathways",
        showCategory = 10)

```


```{r, message=FALSE}
go_cc <- enrichGO(
  gene          = lipid_genes_ensembl,
  OrgDb         = org.Dr.eg.db,
  keyType       = "ENSEMBL",
  ont           = "CC",               # Cellular Component
  pvalueCutoff  = 0.05
)

cnetplot(lipid_go_bp, 
         categorySize = "pvalue", 
         showCategory = 5,
         foldChange = NULL)

```


## Impact on mitochondrial genes 


```{r}
library(biomaRt)

# Connect to Ensembl
mart <- useMart("ensembl", dataset = "drerio_gene_ensembl")

# Retrieve genes with mitochondrial GO terms
mito_genes <- getBM(
  attributes = c("ensembl_gene_id", "external_gene_name", "go_id", "name_1006"),
  filters = "go",
  values = c("GO:0005739", "GO:0005743", "GO:0005759"), 
  mart = mart
)

# Save the list
#write.csv(mito_genes, "zebrafish_mitochondrial_genes.csv", row.names = FALSE)
```

```{r,warning=FALSE}
# Get Ensembl IDs of mitochondrial genes
mito_ensembl_ids <- unique(mito_genes$ensembl_gene_id)


mydata_mito <- subset(mydata, features = mito_ensembl_ids)

mydata_mito <- ScaleData(mydata_mito)
```


```{r}
Idents(mydata_mito) <- "treatment"

# Find DE mitochondrial genes
de_mito <- FindMarkers(
  mydata_mito,
  ident.1 = "befa",
  ident.2 = "gf",
  test.use = "wilcox",  
  logfc.threshold = 0.25,
  min.pct = 0.1
)

sig_mito <- de_mito %>% 
  filter(p_val_adj < 0.1) %>% 
  arrange(desc(abs(avg_log2FC)))

```

```{r, warning=FALSE}
library(dplyr) 

top_mito_genes <- rownames(sig_mito)[1:16]

# Get gene symbols for labels
gene_symbols <- mito_genes[mito_genes$ensembl_gene_id %in% top_mito_genes, 
                          c("ensembl_gene_id", "external_gene_name")]
gene_symbols <- unique(gene_symbols)  

# Create named vector for labeling
gene_labels <- setNames(gene_symbols$external_gene_name, gene_symbols$ensembl_gene_id)


DotPlot(
  mydata_mito,
  features = top_mito_genes,
  group.by = "treatment",
  cols = c("blue", "red"),  
  dot.scale = 6
) +
  scale_x_discrete(labels = gene_labels) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Top DE Mitochondrial Genes: BEFA vs. GF")
```


```{r}

go_results <- enrichGO(
  gene = rownames(sig_mito),
  OrgDb = org.Dr.eg.db,
  keyType = "ENSEMBL",
  ont = "BP",  
  pvalueCutoff = 0.1
)

mito_pathways <- go_results %>% 
  filter(grepl("mitochondr|respir|ATP synth|oxidative", Description, ignore.case = TRUE))

barplot(mito_pathways, 
        showCategory = 10, 
        title = "Enriched Mitochondrial Pathways")


```
